---
title: "Lead IQ Data Report"
author: "Emily Kokesh"
date: "2025-11-23"
output: html_document
---

## Imports

```{r}
knitr::opts_chunk$set(
echo = TRUE,
message = FALSE,
warning = FALSE
)

library(tidyverse)
library(knitr)

```

## Introduction

This report summarizes the relationship between children’s IQ scores and
their proximity to a lead smelter.

```{r}

lead_raw <- read_csv("../DataRaw/lead-iq-01.csv")

# Basic counts used for in-line text

n_total   <- nrow(lead_raw)
n_near    <- sum(lead_raw$Smelter == "Near")
n_far     <- sum(lead_raw$Smelter == "Far")

overall_mean_IQ <- mean(lead_raw$IQ)

```

```{r clean-data}
# Recode the incorrect IQ value of 999 to 99
lead <- lead_raw %>%
  mutate(
    IQ = if_else(IQ == 999, 99L, IQ)
  )

# Recompute counts and means based on the cleaned data
n_total   <- nrow(lead)
n_near    <- sum(lead$Smelter == "Near")
n_far     <- sum(lead$Smelter == "Far")

overall_mean_IQ <- mean(lead$IQ)

```

There are `r n_total` children in this dataset:
`r n_near` lived near the smelter and `r n_far` lived farther away.
The overall mean IQ in the sample is `r round(overall_mean_IQ, 1)`.

## Descriptive statistics table
```{r}
summary_by_smelter <- lead %>%
group_by(Smelter) %>%
summarise(
n       = n(),
mean_IQ = mean(IQ),
sd_IQ   = sd(IQ),
min_IQ  = min(IQ),
max_IQ  = max(IQ),
.groups = "drop"
)


kable(
summary_by_smelter,
digits = 1,
caption = "Descriptive statistics for IQ scores by distance from the smelter."
)

```

The table above shows the number of children, mean IQ, and spread of IQ
scores for those living near versus far from the smelter. At this stage,
the group means are `r round(summary_by_smelter$mean_IQ[summary_by_smelter$Smelter == "Near"], 1)`
for children living near the smelter and
`r round(summary_by_smelter$mean_IQ[summary_by_smelter$Smelter == "Far"], 1)`
for those living farther away.

## IQ by smelter proximity plot
```{r}
ggplot(lead, aes(x = Smelter, y = IQ)) +
geom_boxplot() +
labs(
title = "IQ scores by proximity to lead smelter",
x = "Proximity to smelter",
y = "IQ score (WISC)"
) +
theme_minimal()

```

The boxplot shows the distribution of IQ scores for children living near
and far from the smelter. We can visually compare typical IQ levels and
the variability in each group, as well as identify any potential outliers.

```{r}
#write_csv(lead, "../DataProcessed/lead-iq-01-clean.csv")
```


